---
title: "ring"
author: "Rich FitzJohn"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{ring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

``` {r echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE)
```

This package implements ring buffers.  A ring buffer can be used as
a first-in-first-out (FIFO) buffer where the maximum size is known
ahead of time.  Because they do not grow in size, they are useful
to avoid using more and more memory as a process runs.  Because the
data reading and writing happens in an (apparently) circular way,
once data is added to the buffer it is not copied (in contrast if
you used a vector then every time data is consumed you'd have to
shuffle the vector around).

`ring` implements two different ring buffers that will likely suit
different applications.

* `ring_buffer_bytes` is imlemented as a byte array in C, possibly
  with a "stride" indicating a set of bytes that go together.  Once
  the data reaches the end of the array we start writing to the
  beginning again.
* `ring_buffer_env` is implemented as a doubly linked list using
  R's environments.  This buffer can hold arbitrary R objects at
  each position.

# The environment buffer `ring_buffer_env`

This is the simplest buffer to understand because we don't have to
deal with raw vectors.

To create a buffer that can hold up to 100 elements, use the
`ring_buffer_env` function:
``` {r }
buf <- ring::ring_buffer_env(100)
```

This is an [R6](https://github.com/wch/R6) class:
``` {r }
buf
```

Operations on the class happen by running methods using `$`.  So
the size of the buffer:
``` {r }
buf$size()
```

...the number of elements free and used:
``` {r }
buf$free()
buf$used()
```

...whether the buffer is empty or full:
``` {r }
buf$empty()
buf$full()
```

To start using the buffer we need to put some data in it.  There
are two functions for adding data:

* `buf$set(data, n)` sets `n` elements to be the value `data`
* `buf$push(data, iterate)` pushes `data` into the buffer, with the
  `iterate` argument indicating if we should iterate over `data` or
  treat it as a single element

So to set the first 5 elements to be "a", "b", ..., "e", use:
``` {r }
buf$push(letters[1:5])
```

The buffer is no longer empty
``` {r }
buf$empty()
```

...having 5 elements:
``` {r }
buf$used()
```

...and room for 95 more:
``` {r }
buf$free()
```

To read the content of the buffer without modifying it, use
`read(n)` where `n` is the number of elements to read.  This
*always* returns a list of length `n`:
``` {r }
buf$read(1)
buf$read(2)
```

If you try to read too far, then the buffer will underflow and you
will get an error:
``` {r error=TRUE}
buf$read(20)
```

If you just want the the first element, use `tail_data()`
``` {r }
buf$tail_data()
```

The tail returns the first element in (so the buffer naturally
operates as a first-in-first-out queue).

You can also read the most recently added element with `head_data()`
``` {r }
buf$head_data()
```

And you can offset these by an integer number of steps.  So moving
one position into the buffer from the tail gets the second element
added:
``` {r }
buf$tail_offset_data(1)
```

or moving three elements into the buffer from the head (most
recently added elemnt) gets the same bit of data
``` {r }
buf$head_offset_data(3)
```

The above operations are all nondestructive -- they leave the
buffer unchanged.  To consume elements, use `take(n)` which
operates the same way as `read` but it also moves the buffer
`tail`; it consumes elements leaving space for more.
``` {r }
buf$free()
buf$take(1)
buf$free()
```

Now we have consumed an element the tail has moved along, so
`tail_data` contains "b" and "a" is removed from the buffer:
``` {r }
buf$tail_data()
```

To reset the buffer, use `reset()`.  This empies the buffer of all data:
``` {r }
buf$reset()
buf$used()
buf$empty()
```

## Application: simulation with recent history

The whole point of the ring buffer though is that we can push
things onto it and pull the most recent out, even when the number
of things pushed *overall* is larger than the buffer size.

So imagine a simulation where we need to keep track of the last 5
steps.  The simulation is a random walk.
``` {r }
step <- function(x) {
  if (runif(1) < 0.5) x - 1L else x + 1L
}

x <- 0L
buf <- ring::ring_buffer_env(5)
h <- integer(20)
buf$push(x)
h[1L] <- x

set.seed(1)
for (i in seq_len(length(h) - 1L)) {
  x <- step(x)
  buf$push(x)
  h[i + 1L] <- x
}
```

The whole history:
``` {r }
h
```

The last 5 steps:
``` {r }
unlist(buf$read(5))
```

So we could rewrite the simulation so that the random walk tends up
if the last few steps have been increases and tends down if the
last few steps have been decreases:
``` {r }
step <- function(x) {
  if (length(x) > 1) {
    p <- mean(diff(x)) / 2 + 0.5
  } else {
    p <- 0.5
  }
  if (runif(1) < p) x[length(x)] - 1L else x[length(x)] + 1L
}

x <- 0L
buf <- ring::ring_buffer_env(5)
h <- integer(100)
buf$push(x)
h[1L] <- x

set.seed(1)
for (i in seq_len(length(h) - 1L)) {
  x <- step(unlist(buf$read(buf$used())))
  buf$push(x)
  h[i + 1L] <- x
}
```

Now we have a simulation with a strong mean reverting tendency:
``` {r fig.width=7}
par(mar=c(4, 4, .5, .5))
plot(h, type="l", xlab="step", ylab="y", las=1)
```

Because the buffer always holds the last 5 (or fewer) elements the
book-keeping involved in working with the last few elements out is
simlified.  Ignoring the fact that we hold the entire history in
the fixed size vector `h`, only the last few elements need to be
retained which may be useful if the simulation generates a lot of
data.

A downside of this implementation is that `buf$read()` returns a
list that must be turned into a vector with `unlist`, even though
we know in this case that the simulation will always produce an
integer vector.  The ring buffers described below can help with
that problem.

# The bytes buffer `ring_buffer_bytes`

This is the classical implementation of a ring buffer, and the
implementation is largely due to
[@dhess](https://github.com/dhess/c-ringbuf).

This operates basically the same way as `ring_buffer_env` but with
a few differences:

* The contents of the buffer are raw bytes (R's raw vectors).
  These are a bit fiddly to work with but can be very powerful.
* The `iterate` distinction of `push` disappears because there is
  no ambiguity with R objects
* Reading from the head is not currently implemented.
* There are a few methods that each implementation has that the
  other does not but these are not needed for most uses.  These
  will be documented in the reference documentation.

To construct a buffer of 1000 bytes:
``` {r }
buf <- ring::ring_buffer_bytes(1000)
```

Most of the same methods apply directly:
``` {r }
buf$free()
buf$used()
buf$full()
buf$empty()
```

Generate a byte sequence:
``` {r }
bytes <- as.raw(0:255)
```

...and push them into the buffer:
``` {r }
buf$push(bytes)
```

## Striding

Single bytes can hold only values 0 to 255 (or character
equivalents, such as `a` becomes `r charToRaw("a")` via
`charToRaw("a")`.  But if you want to hold a full integer, that
(usually) takes 4 bytes, a double (usually) takes 8.

To allow this, a bytes buffer can be "strided"; this indicates the
number of consecutive bytes that should together make up one
logical entry.  The buffer then contains `size` of these.  So to
create a buffer of 100 entries, each of `8` bytes you could do:
``` {r }
buf <- ring::ring_buffer_bytes(100, 8)
```

Each element pushed onto the buffer must have the correct size.  So
to push the byte sequence 1..8 onto the buffer:
``` {r }
buf$push(as.raw(1:8))
```

but if you pushed more or less it would be an error:
``` {r error=TRUE}
buf$push(as.raw(1:4))
```

Reading happens in *logical* units, not bytes:
``` {r }
buf$read(1)
```

and you can get the number of elements used:
``` {r }
buf$used()
```

or the number of bytes
``` {r }
buf$used(bytes=TRUE)
```

## The typed bytes buffer `ring_buffer_bytes_typed`

If 8 bytes is a double, it should be possible to make a bytes
buffer that holds one (or more) doubles per entry.  That is what
the `ring_buffer_bytes_typed` buffer does, with a few corner cases
dealt with.  To use, you decide what the *R interpretation* of an
entry is, it will determine the size per entry and appropriate
encoding and decoding functions and you can ignore that it is
storing bytes.  For performance reasons this does not use R's
serialisation and simply copies the data stored in vectors.

For example, to make a buffer of 10 elements, each of which is a
single real number (double), use:
``` {r }
buf <- ring::ring_buffer_bytes_typed(10, double(1))
```

onto which real numbers can be pushed:
``` {r }
buf$push(pi)
```

And retrieve the data.
``` {r }
buf$take(1)
```

Entries can contain more than one number; to make a buffer of
length 10, each element of which is a vector of 5 doubles:
``` {r }
buf <- ring::ring_buffer_bytes_typed(10, double(5))
buf$push(rnorm(5))
buf$read(1)
```

Because this is just implemented as a byte array, we can just push
a bunch of numbers straight into the buffer:
``` {r }
buf$push(rnorm(5 * 10))
```

With elements in the buffer, we can request them.  The integer
argument of `take` indicates the number of groups of 5 doubles we
would like back:
``` {r }
buf$take(1)
```

If you try to take more than is in the buffer it is an error:
``` {r error=TRUE}
buf$take(10)
```

# Possible applications

## Ring vector

The ring buffer above is a better data structure to implement the
simulation than the environment buffer is because the expected
elements in each entry of the buffer are the same.  But with a bit
of S3 syntactic sugar we can do a bit better.
``` {r }
v <- ring::ring_vector(5, "integer", FALSE)
```

Convert back out to be an R vector (involves a copy)
``` {r }
v[]
```

To add things to the vector, use the `push` generic:
``` {r }
ring::push(v, 1L)
v[]
```

This can push multiple items on at once:
``` {r }
ring::push(v, 2:4)
v[]
length(v)
```

Random read access works:
``` {r }
v[3]
v[[1]]
```

Resetting the buffer zeros this all:
``` {r }
v$buf$reset()
length(v)
```

Returning to the simulation example from above:
``` {r }
buf <- ring::ring_vector(5, "integer", FALSE)
h <- integer(20)
x <- 0L
ring::push(buf, x)
h[1L] <- x

step <- function(x) {
  if (runif(1) < 0.5) x - 1L else x + 1L
}

set.seed(1)
for (i in seq_len(length(h) - 1L)) {
  x <- step(x)
  ring::push(buf, x)
  h[i + 1L] <- x
}
```

The whole history:
``` {r }
h
```

The last 5 steps:
``` {r }
buf[]
```

Now, rewriting again, this time with the step function taking the
buffer itself.  This simplifies the implementation, with most of
the details being handled by the S3 methods for `length`, `push`
and `[`.
``` {r }
step <- function(x) {
  if (length(x) > 1) {
    p <- mean(diff(x[])) / 2 + 0.5
  } else {
    p <- 0.5
  }
  if (runif(1) < p) x[length(x)] - 1L else x[length(x)] + 1L
}

buf <- ring::ring_vector(5, "integer", FALSE)
h <- integer(100)
x <- 0L

ring::push(buf, x)
h[1L] <- x

set.seed(1)
for (i in seq_len(length(h) - 1L)) {
  x <- step(buf)
  ring::push(buf, x)
  h[i + 1L] <- x
}

par(mar=c(4, 4, .5, .5))
plot(h, type="l", xlab="step", ylab="y", las=1)
```

## Ring matrix with `ring_matrix`

The `ring_matrix` data structure generalises the `ring_vector`; it
is a buffer that looks to R like a matrix that grows by adding rows
at the bottom and shrinks by consuming rows at the top.

This is even more contrived than above, but consider simultaneously
simulating the movement of `n` random particles with the same
reflecting random walk as above:

``` {r }
n <- 10
m <- ring::ring_matrix(5, n, "integer", FALSE)
```

The current state of the matrix is:
``` {r }
m[]
```

We can set the initial state as:
``` {r }
ring::push(m, matrix(0L, 1, n))
m[]

step <- function(x) {
  if (nrow(x) > 1) {
    p <- colMeans(diff(m[])) / 2 + 0.5
  } else {
    p <- rep(0.5, ncol(m))
  }
  x[nrow(x), ] + as.integer(ifelse(runif(length(p)) < p, -1, 1L))
}

m <- ring::ring_matrix(5, n, "integer", FALSE)
x <- rep(0L, n)
ring::push(m, x)

h <- matrix(NA, 200, n)
h[1, ] <- x
set.seed(1)
for (i in seq_len(nrow(h) - 1L)) {
  x <- step(m)
  ring::push(m, x)
  h[i + 1L, ] <- x
}

par(mar=c(4, 4, .5, .5))
matplot(h, type="l", lty=1)
```
